Experimentally informed site-specific substitution models deepen phylogenetic estimates of the divergence of viral lineages

### 2018-02-26 manuscript/manuscript.tex
small tweaks to simulations figure
added citation to model feature section

Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
Codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
The relative rate of these substitutions is referred to as dN/dS or $\omega$.
In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called M0.
We will refer to the M0 variant of the \citet{goldman1994codon} model simply as GY94.
The gene-wide $\omega$ is usually $<1$\jdbcomment{add citation, most of the site-specific $\omega$ papers will talk about how gene-wide is not sensitive, so we can cite some of those}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.

@@

\subsection*{Simulations demonstrate how failure to model site-specific stationary states leads to branch-length underestimation.}
\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
\caption{\label{fig:simulations}
\textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
We simulated alignments along a phylogenetic tree of HA genes (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences measured by deep mutational scanning of an H1 HA~\citep{doud2016accurate}.
We then inferred the branch lengths of this tree on the simulated alignments.
The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.
We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
The grey dashed line at $y=x$ represents the behavior of an unbiased estimator. 
\jdbcomment{Because of order we will discuss in text, switch order of ExpCM and ExpCM averaged across sites.}
}
\end{figure}
To directly demonstrate the effect of stationary state and $\Gamma\omega$ rate variation on branch-length estimation, we tested the ability of a variety of models to accurately infer branch lengths on simulated data (\ref{fig:simulations}).
Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees}) using ExpCM parameterized by the actual amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}. We then estimated the branch lengths from the simulated sequences using all the substitution models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths used in the simulations.

>>>>

Here we consider how purifying selection is handled by codon models, which are the most accurate of the three classes (nucleotide, amino acid, and nucleotide) of phylogenetic substitution models in widespread use~\citep{arenas2015trends}.
Codon models distinguish between two types of substitutions: synonymous and nonsynonymous.
The relative rate of these substitutions is referred to as dN/dS or $\omega$.
In their simplest form, codon substitution models fit a single $\omega$ that represents the gene-wide average selection on nonsynonymous mutations relative to synonymous ones.
Here we will use such substitution models in the form proposed by \citet{goldman1994codon}.
Using the classification scheme of \citet{yang2000codon}, when these models have a single gene-wide $\omega$ they are called M0.
We will refer to the M0 variant of the \citet{goldman1994codon} model simply as GY94.
The gene-wide $\omega$ is usually $<1$\citep{murrell2015gene}, and crudely represents the fact that many amino-acid substitutions are under purifying selection.

@@

\subsection*{Simulations demonstrate how failure to model site-specific stationary states leads to branch-length underestimation.}
\begin{figure}
\centerline{\includegraphics[width=0.85\textwidth]{figures/simulations}}
\caption{\label{fig:simulations}
\textbf{Branch lengths inferred on data simulated under a model with site-specific amino-acid preferences.} 
We simulated alignments along a phylogenetic tree of HA genes (see Figure~\ref{fig:empirical_trees}) using an ExpCM parameterized by the actual site-specific amino-acid preferences measured by deep mutational scanning of an H1 HA~\citep{doud2016accurate}.
We then inferred the branch lengths of this tree on the simulated alignments.
The inferred branch lengths for various models are plotted on the x-axis, and the actual branch lengths used in the simulations are on the y-axis.
We performed 10 simulations and inferences, and gray points show each inferred branch length from each simulation, and black points show the average of each branch length across simulations.
The grey dashed line at $y=x$ represents the behavior of an unbiased estimator. 
}
\end{figure}
To directly demonstrate the effect of stationary state and $\Gamma\omega$ rate variation on branch-length estimation, we tested the ability of a variety of models to accurately infer branch lengths on simulated data (\ref{fig:simulations}).
Specifically, we simulated alignments of sequences along the HA phylogenetic tree in \ref{fig:empirical_trees}) using ExpCM parameterized by the actual amino-acid preferences of H1 HA as experimentally measured by deep mutational scanning~\citep{doud2016accurate}. We then estimated the branch lengths from the simulated sequences using all the substitution models in \ref{fig:model_feature}C, and compared these estimates to the actual branch lengths used in the simulations.
